---
title: "Assignment on Assignment Report on Mapping Crop-types using Available Satellite
  RS Data and VGI In-situ Data in R"
author: "Pranish Shrestha"
date: "2023-11-02"
output:
  html_document: default
  editor_options: 
  markdown: 
    wrap: 72
---
```{r include=FALSE}
# Load the required R packages
library(sf)      # For working with shapefiles
library(leaflet) # For geospatial mapping
library(rgdal)
library(raster)
library(terra)
library(dplyr)
library(ggplot2)
library(rgeos)
library(viridis)
library(rasterVis)
library(randomForest)
library(leaflet.extras)
library(knitr)
library(tidyr)
#Load the in-situ data (shapefile)
insitu_shp <- st_read("LULC_Crop-Types_In-SituData2021_USP/LULC_Crop-Types_In-SituData2021_USP.shp")
# View the first few rows of the attribute data
head(insitu_shp)

# List all the column names in shapefile
column_names <- names(insitu_shp)
print(column_names)

# Remove the non-usable columns from  data frame
insitu_shp <- insitu_shp %>%
  select(-data_Other, -data_Photo, -data_UserN, -data_Phone, -data_Subsc, -data_Email, -data_EndTi, -data_meta_, -F17, -F18)

# See the unique Crop Types
unique_data_CropT <- unique(insitu_shp$data_CropT)
print(unique_data_CropT)

# See the unique Landuse  Types
unique_data_LULC <- unique(insitu_shp$data_LULC)
print(unique_data_LULC)

insitu_shp <- insitu_shp %>%
  mutate(NewColumn = ifelse(data_LULC == "Agriculture", data_CropT, data_LULC))

# View the first few rows of the updated data frame
head(insitu_shp)


insitu_shp <- insitu_shp %>%
  mutate(LU_CT = ifelse(data_LULC %in% c("Residential", "WaterBodies", "PublicUse", "CULARCH", "Forest", "Commercial", "Industrial"), "Mixed_Area", data_LULC))

# View the first few rows of the updated data frame
head(insitu_shp)

insitu_shp <- insitu_shp %>%
  mutate(LU_CT = ifelse(NewColumn %in% c("Residential", "WaterBodies", "PublicUse", "CULARCH", "Forest", "Commercial", "Industrial"), "Mixed_Area", NewColumn))

# View the first few rows of the updated data frame
head(insitu_shp)

# Again Remove the non-usable columns from  data frame
insitu_shp <- insitu_shp %>%
  select(-ClassType, -data_CropT, -data_LULC, -NewColumn)
head(insitu_shp)






# Load the district shapefile
district_shp <- st_read("nepal_data/hermes_NPL_new_wgs_2.shp")

# Define the target districts
target_districts <- c("Dhanusha", "Mahottari", "Sarlahi")

# Filter the shapefile to include only the target districts
filtered_districts <- district_shp[district_shp$DISTRICT %in% target_districts, ]

# Define a custom color palette for the specific class types
color_palette <- colorFactor(
  palette = c("lightgreen", "yellow", "darkgreen", "cyan", "orange","purple"),  # Define colors for each class type
  domain = c("Sugarcane", "PaddyRice", "Orchid","Bamboo", "OtherCrop", "Mixed_Area")  # Define class types
)

# Create a leaflet map with OpenStreetMap (OSM) as the base map
leaflet() %>%
  addProviderTiles(providers$OpenStreetMap) %>%
  addPolygons(data = filtered_districts, color = "black", weight = 1, opacity = 1, fillOpacity = 0.2) %>%
  addCircleMarkers(data = insitu_shp,
                   lng = ~st_coordinates(insitu_shp$geometry)[, 1],
                   lat = ~st_coordinates(insitu_shp$geometry)[, 2],
                   popup = ~LU_CT,
                   color = ~color_palette(LU_CT),
                   radius = 1)%>%
  setView(lng = 85.793, lat = 26.953, zoom = 8.8)




#importing sentinel image data

red_band <- raster("T45RUK_20211223T045221_B04.jp2")
nir_band <- raster("T45RUK_20211223T045221_B08.jp2")
TCI <- raster("T45RUK_20211223T045221_TCI.jp2")






# Calculation NDVI
ndvi_funtion <- function(red_band, nir_band) {
  ndvi <- (nir_band - red_band) / (nir_band + red_band)
  return(ndvi)
}

ndvi <- ndvi_funtion(red_band , nir_band)

#plotting NDVI
gplot(ndvi) +
  geom_raster(aes(x = x, y = y, fill = value)) +
  coord_quickmap() +
  ggtitle("NDVI") +
  xlab("Easting") +
  ylab("Northing") +
  theme_classic() +   					   
  theme(plot.title = element_text(hjust = 0.5),             
        text = element_text(size=15),		       	    
        axis.text.x = element_text(angle = 90, hjust = 1)) 


# Check the CRS ofNDVI data
crs_ndvi <- st_crs(ndvi)
# Print the CRS
print(crs_ndvi)


# Check the CRS of shapefile
crs_insitu <- st_crs(insitu_shp)
# Print the CRS
print(crs_insitu)


crs_target <- st_crs("+init=EPSG:32645")

insitu_projected <- st_transform(insitu_shp, crs_target)


# Extract NDVI values for insitu data locations
ndvi_values <- raster::extract(ndvi, insitu_projected)

# Add the extracted NDVI values to the insitu_filtered_projected data
insitu_combined <- cbind(insitu_projected, NDVI = ndvi_values)

head(insitu_combined)

nrow(insitu_combined)

# Remove rows with NA values in the NDVI column
insitu_combined <- insitu_combined[!is.na(insitu_combined$NDVI), ]
nrow(insitu_combined)


# Define the proportion for the training data (70% for training, 30% for testing)
train_prop <- 0.7

# Set a seed for reproducibility
set.seed(123)

# Initialize empty data frames for training and testing sets
train_data <- data.frame()
test_data <- data.frame()

# Define class names
class_names <- c("Bamboo", "Orchid", "PaddyRice", "Sugarcane", "OtherCrop", "Mixed_Area")

# Split the data into training and testing sets with at least one sample from each class
for (class_name in class_names) {
  # Filter rows for the current class
  class_data <- insitu_combined[insitu_combined$LU_CT == class_name, ]
  
  # Calculate the number of samples for training
  num_train_samples <- round(train_prop * nrow(class_data))
  
  # Randomly select samples for training and the rest for testing
  class_train_index <- sample(1:nrow(class_data), num_train_samples)
  class_test_index <- setdiff(1:nrow(class_data), class_train_index)
  
  # Add the selected samples to the training and testing sets
  train_data <- rbind(train_data, class_data[class_train_index, ])
  test_data <- rbind(test_data, class_data[class_test_index, ])
}

# Check the composition of the training and testing sets
table(train_data$LU_CT)
table(test_data$LU_CT)

# Train the random forest model
train_data$LU_CT <- as.factor(train_data$LU_CT)
rf_model <- randomForest(LU_CT ~ NDVI, data = train_data, ntree = 100)

# Make predictions on the test data
rf_predictions <- predict(rf_model, test_data)

predicted_classes <- predict(rf_model, test_data)
actual_classes <- test_data$LU_CT
confusion_matrix <- table(Predicted = predicted_classes, Actual = actual_classes)
print(confusion_matrix)


# Convert the confusion matrix to a data frame
confusion_matrix_df <- as.data.frame(as.table(confusion_matrix))

# Rename the columns for better labels
colnames(confusion_matrix_df) <- c("Predicted", "Actual", "Frequency")

# Create the heatmap using ggplot2
heatmap_plot <- ggplot(confusion_matrix_df, aes(Actual, Predicted, fill = Frequency)) +
  geom_tile() +
  geom_text(aes(label = Frequency), vjust = 1, size = 4) +
  scale_fill_gradient(low = "pink", high = "red") +  # Define color palette
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels
  labs(title = "Confusion Matrix Heatmap", x = "Actual", y = "Predicted")

# Display the heatmap
print(heatmap_plot)

# Calculate accuracy
accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
cat("Accuracy: ", round(accuracy, 2), "\n")


# Initialize vectors to store precision, recall, and F1 values
precision_values <- numeric(length(class_names))
recall_values <- numeric(length(class_names))
f1_values <- numeric(length(class_names))


for (i in 1:length(class_names)) {
  class_name <- class_names[i]
  TP <- confusion_matrix[class_name, class_name]
  FP <- sum(confusion_matrix[, class_name]) - TP
  FN <- sum(confusion_matrix[class_name, ]) - TP
  precision <- TP / (TP + FP)
  recall <- TP / (TP + FN)
  f1_value <- 2 * (precision * recall) / (precision + recall)
  
 
  
  # Store values in respective vectors
  precision_values[i] <- precision
  recall_values[i] <- recall
  f1_values[i] <- f1_value
}

# Create a data frame with  results
results_df <- data.frame(
  Class = class_names,
  Precision = round(precision_values, 2),
  Recall = round(recall_values, 2),
  F1_Score = round(f1_values, 2)
)

# Use kable to format the results as a table
kable(results_df, caption = "Precision, Recall, and F1-Score for Each Class")

# Transform the data for a grouped bar chart
results_long <- results_df %>%
  pivot_longer(cols = c(Precision, Recall, F1_Score), names_to = "Metric", values_to = "Value")

# Create a grouped bar chart for Precision, Recall, and F1-Score
grouped_bar <- ggplot(results_long, aes(Class, Value, fill = Metric)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Performance Metrics for Each Class", x = "Class", y = "Value") +
  scale_fill_manual(values = c("Precision" = "blue", "Recall" = "green", "F1_Score" = "red")) +
  theme_minimal() +
  theme(legend.title = element_blank())

# Display the grouped bar chart
print(grouped_bar)

# Initialize vectors to store producer accuracy (PA) and user accuracy (UA)
PA_values <- numeric(length(class_names))
UA_values <- numeric(length(class_names))


for (i in 1:length(class_names)) {
  class_name <- class_names[i]
  
  # Calculate Producer Accuracy (PA)
  PA <- confusion_matrix[class_name, class_name] / sum(confusion_matrix[class_name, ])
  
  # Calculate User Accuracy (UA)
  UA <- confusion_matrix[class_name, class_name] / sum(confusion_matrix[, class_name])
  

  
  # Store values in respective vectors
  PA_values[i] <- PA
  UA_values[i] <- UA
}

# Create a data frame for  Producer Accuracy and User Accuracy results
PA_UA_df <- data.frame(
  Class = class_names,
  Producer_Accuracy = round(PA_values, 2),
  User_Accuracy = round(UA_values, 2)
)

# Use kable to format the results as a table
kable(PA_UA_df, caption = "Producer and User Accuracy for Each Class")

# Transform the data for a grouped bar chart for PA and UA
PA_UA_long <- PA_UA_df %>%
  pivot_longer(cols = c(Producer_Accuracy, User_Accuracy), names_to = "Metric", values_to = "Value")

# Create a grouped bar chart for PA and UA
grouped_bar_PA_UA <- ggplot(PA_UA_long, aes(Class, Value, fill = Metric)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Producer and User Accuracy for Each Class", x = "Class", y = "Value") +
  scale_fill_manual(values = c("Producer_Accuracy" = "orange", "User_Accuracy" = "maroon")) +
  theme_minimal() +
  theme(legend.title = element_blank())

# Display the grouped bar chart for PA and UA
print(grouped_bar_PA_UA)

```

# Executive Summary

In this report, We have conducted an analysis of mapping crop types using
Sentinel satellite image data and volunteered geographic information
(VGI) in R. We have used R, along with various packages, for geospatial
analysis and machine learning to achieve the objectives. The primary
objective was to classify different crop types based on the analysis
using NDVI and in-situ observations.

**Key Findings:**

1.  **Data Preparation**: We loaded in-situ data and have several
    proprocessing tasks such as cleaning, removing unusable column, and
    merging columns as required to represent land use and crop types.

2.  **Data Visualization**: We used Leaflet to create an interactive map
    that displays the in-situ data points over OSM Map, categorized by
    land use and cover types with various color pallete to represent the
    data.

3.  **Satellite Data**: We imported and processed Sentinel satellite
    data directly from web broswer and calculated the Normalized
    Difference Vegetation Index (NDVI).

4.  **NDVI Analysis**: We calculated the NDVI values from the satellite
    data and created a heatmap to visualize vegetation health and land
    cover patterns.

5.  **Machine Learning**: We implemented a random forest classifier
    model to predict land cover classes using NDVI as a predictor
    variable. The model was trained and tested using in-situ data in the
    ratio of 70%:30%.

6.  **Model Evaluation**: We evaluated the model's performance using a
    confusion matrix and calculated accuracy, precision, recall, and
    F1-score for each land cover class. We also determined producer and
    user accuracy values.

**Recommendations:**

Based on the project findings, the following recommendations can be
made:

-   The more insitu data should be collect with more accuratly for
    refining the predictive model.

-   Multiple indices should be used with addition of NDVI to have better
    predictions.

-   Consider conducting field surveys and data collection to further
    validate the model's predictions and enhance its reliability.

-   Extend the analysis to a larger geographical area to assess land use
    and land cover patterns more widely

-   Explore with using more machine learning techniques to compare the
    results to have better reliability.

# Table of Contents

-   [Introduction](#introduction)
-   [Methods](#methods)
-   [Results](#results)
-   [Discussion](#discussion)
-   [Conclusion](#conclusion)
-   [References](#references)

# Introduction {#introduction}



The effective management and monitoring of land use data is very crucial
for urban planning as well as the ensuring the agricultural production
with environment sustainability. With the accurate information regarding
crop types and land use we chan have better decision-making in
agriculture, resource allocation, and land planning. In this project,we
have the utilized sentinel satellite remote sensing data and volunteered
geographic information (VGI) to have Random Forest Classifier Model and
predicted classsficiation of crop types and buildup areas

**Context**

This report focuses on a specific case study involving the mapping of
crop types using available satellite remote sensing data and VGI data in
the context of Madesh Province of Nepal specially Sarlahi, Dhanusha and
Mohattari District. The in-situ data has been collected after the end of
Mansoon in the month of October to Janurary. The sample size is approx
600, and is backed with photographic proof. The study area has been
chosen due to its significance in agriculture and the presence of
diverse crop types. The regions is known epicenter of for crop
production for the country

### Objectives

The primary objectives of this report are as follows:

-   To analyze the feasibility and effectiveness of classifying crop
    types using satellite remote sensing data and VGI.

-   To visualize the insitu data over the map

-   To assess the accuracy and performance of the classification model
    with confusion matrix, overall accuracy and individual class
    performance.

-   To analyse benifits and limitations of the approach and
    recommendations for further improvements in land use mapping.

# Methods

## ***Data Collection:***

**Satellite Remote Sensing Data:** The first component of our data
collection involved obtaining satellite remote sensing data. We acquired
multi-spectral Sentinel-2 imagery from
<https://dataspace.copernicus.eu/> covering the study area during the
month of Novermber, 2021. Sentinel-2 is part of the European Space
Agency's Copernicus program and provides high-resolution, multispectral
imagery of the Earth's surface. We selected an data of November, 2021
that aligns with our study objectives as our insitu data was on same
time frame. The imagery consisted of various spectral bands, including
red (B04) and near-infrared (B08), which are essential for calculating
the normalized difference vegetation index (NDVI).

**Volunteered Geographic Information (VGI):** In parallel with satellite
data, we had availed in-situ VGI data. The in-situ data used was
collected after the end of the monsoon from October 23, 2021, to January
6, 2021. Participants provided valuable ground-level information about
crop types in the study are with photograph and geographical coordinates
associated with each observation.

### Insitu Data Visualization

```{r echo=FALSE}
leaflet() %>%
  addProviderTiles(providers$OpenStreetMap) %>%
  addPolygons(data = filtered_districts, color = "black", weight = 1, opacity = 1, fillOpacity = 0.2) %>%
  addCircleMarkers(data = insitu_shp,
                   lng = ~st_coordinates(insitu_shp$geometry)[, 1],
                   lat = ~st_coordinates(insitu_shp$geometry)[, 2],
                   popup = ~LU_CT,
                   color = ~color_palette(LU_CT),
                   radius = 1)%>%
  setView(lng = 85.793, lat = 26.953, zoom = 8.8)
```

## Analysis 
In this section, we will describe the methods and techniques
used to process and analyze the collected remote sensing and in-situ
data. Our primary goal is to classify land use and crop types in the
study area by integrating satellite imagery with in-situ data.

**Libraries:**

-   **R Language:** We conducted our data analysis and geospatial
    processing using the R programming language. R is a powerful tool
    for statistical analysis and geospatial data handling, making it
    well-suited for our research objectives.

-   **R Packages:** Several R packages were essential for our analysis.

    -   **sf, terra and raster:** These packages allowed us to work with
        geospatial data, including shapefiles and raster imagery. We
        calculated the Normalized Difference Vegetation Index (NDVI)
        using these packages to assess vegetation health.

    -   **dplyr:** This package was used for data manipulation, allowing
        us to filter, mutate, and aggregate datasets effectively.

    -   **viridis:** We utilized the viridis package for defining custom
        color palettes in our visualizations, making it easier to
        distinguish between land cover classes.

    -   **randomForest:** We employed the random forest algorithm for
        supervised classification tasks.

    -   **knitr,tidyr,ggplot2:** These packages were used for data
        visualization, including the creation of plots, table and
        heatmaps.

    -   **leaflet,leaflet.extras:** For the interactive mapping
        component of our analysis, we utilized the leaflet package.

### ***Data Preprocessing:*** 

The data was clean, check for unusable columns and removed it. The new
column LU_CT was created to store crop types and other build up area
merges as Mixed_area.

**Data Projection and Alignment:** To ensure data compatibility, the
satellite remote sensing data and VGI were projected and aligned to a
common coordinate reference system (CRS, EPSG:32645).

**NDVI Calculation:** The NDVI was computed using the red and
near-infrared bands of the satellite imagery. The NDVI calculation was
performed using the formula (NIR - Red) / (NIR + Red), where NIR
represents the near-infrared reflectance values and Red represents the
red band reflectance values. This index is very helpful for
distinguishing vegetation health and land use. \### NDVI Visualitzation

```{r echo=FALSE}
gplot(ndvi) +
  geom_raster(aes(x = x, y = y, fill = value)) +
  coord_quickmap() +
  ggtitle("NDVI") +
  xlab("Easting") +
  ylab("Northing") +
  theme_classic() +   					   
  theme(plot.title = element_text(hjust = 0.5),             
        text = element_text(size=15),		       	    
        axis.text.x = element_text(angle = 90, hjust = 1)) 
```

**Data Splitting** The dataset was divided into training and testing
sets in a proportion of 70% for training and 30% for testing. At first
the combined dataset of ndvi and insitu data was cleaned by removing all
nans. The splitting was done such that that each land cover class was
represented in both the training and testing datasets. A balance between
model learning and evaluation was maintained by preserving a diverse
representation of land cover classes in the test set and train set.

**Classification:**

We employed the Random Forest classification algorithm to categorize
land use and crop types. We trained the classification model using the
NDVI values derived from the satellite data and the reference data set
created from VGI. The Random Forest algorithm is capable in
classification tasks. Our independent variables included the NDVI values
derived from the satellite imagery.

# Results {#results}

## **Classification Results:**

The primary objective of this study was to classify land use and crop
types using the integrated remote sensing and in-situ data. The Random
Forest classification model achieved the following results:

**Overall Accuracy:** The overall accuracy of the classification model
was approximately 47.10%.

**Confusion Matrix:** The confusion matrix below provides detailed
information on the model's performance. It displays the number of
correctly and incorrectly classified samples for each class. It shows
how well our model correctly classified different land types and helps
us calculate important performance metrics like accuracy, precision,
recall, and F1-score. This matrix enables us to assess both the model's
overall performance and its ability to correctly identify specific land
use types.

```{r echo=FALSE}
print(confusion_matrix)
```


**Heatmap Visualization of Confusion Matrix:**

To evaluate the performance of our classification model, we created a
heatmap of the confusion matrix. This heatmap visually represents the
accuracy and misclassification of different land use and crop types.

```{r echo=FALSE}
print(heatmap_plot)
```


**Accuracy Assessment:**

We assessed the accuracy of the classification model by calculating
metrics such as overall accuracy, precision, recall, and F1-score for
each land use and crop class. These metrics provide insights into the
model's performance.

**Precision, Recall, and F1-Score:** Precision, recall, and F1-score
were calculated for each land use and crop class. These metrics provide
insights into the model's performance for individual classes:

-   **Precision:** It measures the accuracy of our model's positive
    predictions. In our project, precision tells us how many of the
    predicted land use or land cover classifications are correct. A high
    precision score indicates a low rate of false positive predictions,
    which is crucial for accurate mapping.

-   **Recall:** This metric assesses how well our model captures all the
    actual positive cases. In our case, it indicates how effectively our
    model identifies the various land use and land cover types present
    in the study area. High recall suggests that the model doesn't miss
    many true land cover categories.

-   **F1-Score:** The F1-Score is the harmonic mean of precision and
    recall. It provides a balanced assessment of our model's
    performance, considering both false positives and false negatives.
    In our project, a high F1-Score is indicative of a balance model
    that accurately predicts land use and land cover categories without
    overly emphasizing precision or recall



```{r echo=FALSE}
kable(results_df, caption = "Precision, Recall, and F1-Score for Each Class")
```
**Bar Chart For Precision, recall and F1 Score**
```{r echo=FALSE}
print(grouped_bar)
```


**Producer Accuracy and User Accuracy:**Producer Accuracy and User
Accuracy were calculated for each land use and crop class. These metrics
help assess the model's performance in terms of omission and commission
errors which have varying implications for different applications. High
UA values indicate that the model's predictions align well with actual
land cover types, providing confidence in the model's outputs.These
metrics are as follows.

```{r echo=FALSE}
kable(PA_UA_df, caption = "Producer and User Accuracy for Each Class")
```


**Bar Chart for Producer and User Accuracy:**

```{r echo=FALSE}
print(grouped_bar_PA_UA)
```


# Discussion {#discussion}

The Random Forest model exhibited an accuracy of approximately 0.47,
indicating that it correctly classified nearly half of the land use and
land cover categories. The classification model demonstrates variable
performance across different land use and crop classes. For instance,
the "PaddyRice" class exhibits relatively higher precision, recall, and
F1-score compared to other classes, indicating better classification
results.

The "Sugarcane" class has a lower precision and recall, suggesting
challenges in distinguishing this class accurately from others.

The overall accuracy of approximately 47.10% indicates a moderate
classification performance, considering the complexity of land use and
crop classification.

The confusion matrix provides valuable insights into the model's
performance, highlighting areas of improvement for specific land use
classes.

# Conclusion {#conclusion}

This project highlights the effectiveness of using Random Forest machine
learning models for land use and land cover classification. It
demonstrates that integrating in-situ data with satellite imagery can
significantly enhance classification accuracy. The model exhibited
niceresults and also identifies areas for future planning. By continuing
to refine the model and incorporating more extensive datasets, we can
further explore the potential of remote sensing and machine learning for
applications in environmental and urban areas. This work lays the
foundation for future research and applications in land use and land
cover classification.

Crop classification requires a large amount of
high-quality in-situ reference data which is very difficult and time-consuming to collect even in
normal conditions and almost impossible to gather
by the experts during a lockdown and transport
bans. However, local youths can fulfill the
demand when they are equipped with the
appropriate tools. In this study, the local youths
collected quality in-situ reference data to classify
freely available high-resolution satellite imageries
such as Sentinel 2 for delineating crop
classes. Voluntarily collected data by youths can
be an inexpensive and easily accessible source of
reference data for crop classification, even in case
of transport restrictions. The approach, however,
requires in-depth training to the potential
volunteers and near real-time feedback for
improved quality in-situ data and capturing
accompanied photographs to support the data
verification.

# References {#references}

1. Volunteered In-Situ Data for Agriculture Crop Mapping: A Case Study of
Nepal By U. S. Panday , A. K. Pratihast , J. Aryal , R. Kayastha 

2. Pratihast, A. K., DeVries, B., Avitabile, V., de
Bruin, S., Kooistra, L., Tekle, M., &
Herold, M. (2014). Combining satellite data
and community-based observations for
forest monitoring. Forests.
https://doi.org/10.3390/f5102464

3. Pratihast, A. K., Herold, M., Avitabile, V., de
Bruin, S., Bartholomeus, H., Souza, C. M.,
& Ribbe, L. (2013). Mobile devices for
community-based REDD+ monitoring: A
case study for central Vietnam. Sensors
(Switzerland), 13, 21–38.
https://doi.org/10.3390/s130100021



# R code chunks:

## Loading Required Libraries

```{r}

# Load the required R packages
library(sf)      # For working with shapefiles
library(leaflet) # For geospatial mapping
library(rgdal)
library(raster)
library(terra)
library(dplyr)
library(ggplot2)
library(rgeos)
library(viridis)
library(rasterVis)
library(randomForest)
library(leaflet.extras)
library(knitr)
library(tidyr)
```

## Loading Available insitu shp file and perform preprocessing task

### Following tasks has been performed for preprocessing task in the insitu data

-   Reviewed the first few rows of the attribute data.
-   Listed all the column names in the shapefile.
-   Removed non-usable columns from the data frame.
-   Explored unique Crop Types and unique Landuse Types in the dataset.
-   Created a new column, "NewColumn," with a conditional statement that
    used "Agriculture" types from "data_CropT" and "data_LULC" values
    such that it stores crop types in place of agriculture.
-   Created another column, "LU_CT," based on a another conditional
    statement that categorized all other land use types into
    "Mixed_Area" but retaining the crop types values.
-   Removed non-usable columns once again from the data frame.

```{r}
#Load the in-situ data (shapefile)
insitu_shp <- st_read("LULC_Crop-Types_In-SituData2021_USP/LULC_Crop-Types_In-SituData2021_USP.shp")
# View the first few rows of the attribute data
head(insitu_shp)

# List all the column names in shapefile
column_names <- names(insitu_shp)
print(column_names)

# Remove the non-usable columns from  data frame
insitu_shp <- insitu_shp %>%
  select(-data_Other, -data_Photo, -data_UserN, -data_Phone, -data_Subsc, -data_Email, -data_EndTi, -data_meta_, -F17, -F18)

# See the unique Crop Types
unique_data_CropT <- unique(insitu_shp$data_CropT)
print(unique_data_CropT)

# See the unique Landuse  Types
unique_data_LULC <- unique(insitu_shp$data_LULC)
print(unique_data_LULC)

insitu_shp <- insitu_shp %>%
  mutate(NewColumn = ifelse(data_LULC == "Agriculture", data_CropT, data_LULC))

# View the first few rows of the updated data frame
head(insitu_shp)


insitu_shp <- insitu_shp %>%
  mutate(LU_CT = ifelse(data_LULC %in% c("Residential", "WaterBodies", "PublicUse", "CULARCH", "Forest", "Commercial", "Industrial"), "Mixed_Area", data_LULC))

# View the first few rows of the updated data frame
head(insitu_shp)

insitu_shp <- insitu_shp %>%
  mutate(LU_CT = ifelse(NewColumn %in% c("Residential", "WaterBodies", "PublicUse", "CULARCH", "Forest", "Commercial", "Industrial"), "Mixed_Area", NewColumn))

# View the first few rows of the updated data frame
head(insitu_shp)

# Again Remove the non-usable columns from  data frame
insitu_shp <- insitu_shp %>%
  select(-ClassType, -data_CropT, -data_LULC, -NewColumn)
head(insitu_shp)
```

### Visualising Area of Availabe insitu data on OSM Map

#### Our area of interest is on Sarlai, Dhanusha and Mahottari District.

A color palette has been created and plotted the data on the OSM map.
Following color is assigned to each of the crop types.

-   Sugarcane : Light Green
-   PaddyRice : Yellow
-   Orchid : Dark Green
-   Bamboo : Cyan
-   OtherCrop : Orange
-   Mixed_Area : Purple

```{r}

# Load the district shapefile
district_shp <- st_read("nepal_data/hermes_NPL_new_wgs_2.shp")

# Define the target districts
target_districts <- c("Dhanusha", "Mahottari", "Sarlahi")

# Filter the shapefile to include only the target districts
filtered_districts <- district_shp[district_shp$DISTRICT %in% target_districts, ]

# Define a custom color palette for the specific class types
color_palette <- colorFactor(
  palette = c("lightgreen", "yellow", "darkgreen", "cyan", "orange","purple"),  # Define colors for each class type
  domain = c("Sugarcane", "PaddyRice", "Orchid","Bamboo", "OtherCrop", "Mixed_Area")  # Define class types
)

# Create a leaflet map with OpenStreetMap (OSM) as the base map
leaflet() %>%
  addProviderTiles(providers$OpenStreetMap) %>%
  addPolygons(data = filtered_districts, color = "black", weight = 1, opacity = 1, fillOpacity = 0.2) %>%
  addCircleMarkers(data = insitu_shp,
                   lng = ~st_coordinates(insitu_shp$geometry)[, 1],
                   lat = ~st_coordinates(insitu_shp$geometry)[, 2],
                   popup = ~LU_CT,
                   color = ~color_palette(LU_CT),
                   radius = 1)%>%
  setView(lng = 85.793, lat = 26.953, zoom = 8.8)
```

### Satellite Data Collection and NDVI Calalculation

For this project Sentinel image has been downloaded. I have downloaded directly
from the web browser and imported. Then NDVI is calculated and the NDVI
image is with ggplot.

```{r}

#importing sentinel image data

red_band <- raster("T45RUK_20211223T045221_B04.jp2")
nir_band <- raster("T45RUK_20211223T045221_B08.jp2")
TCI <- raster("T45RUK_20211223T045221_TCI.jp2")

# Calculation NDVI
ndvi_funtion <- function(red_band, nir_band) {
  ndvi <- (nir_band - red_band) / (nir_band + red_band)
  return(ndvi)
}

ndvi <- ndvi_funtion(red_band , nir_band)

#plotting NDVI
gplot(ndvi) +
  geom_raster(aes(x = x, y = y, fill = value)) +
  coord_quickmap() +
  ggtitle("NDVI") +
  xlab("Easting") +
  ylab("Northing") +
  theme_classic() +   					   
  theme(plot.title = element_text(hjust = 0.5),             
        text = element_text(size=15),		       	    
        axis.text.x = element_text(angle = 90, hjust = 1)) 
```

### Data processing with spatial alignment

#### On this step followsing task has been done 

-   Checked and confirmed the Coordinate Reference System (CRS) of the
    NDVI data.
-   Checked and confirmed the CRS of the in-situ shapefile.
-   Defined the target CRS, which is EPSG:32645.
-   Projected the in-situ data to the target CRS using the st_transform
    function.
-   Extracted NDVI values for the locations specified in the projected
    in-situ data.
-   Created a new data frame named insitu_combined by combining the
    projected in-situ data and the extracted NDVI values.
-   Checked the first few rows of the insitu_combined data frame.
-   Removed rows with NA (Not Available) values in the NDVI column so
    that the area beyond the used satellite image can be avoided.
-   Checked the number of rows in the insitu_combined data frame after
    removing NA values.

```{r}
# Check the CRS ofNDVI data
crs_ndvi <- st_crs(ndvi)
# Print the CRS
print(crs_ndvi)


# Check the CRS of shapefile
crs_insitu <- st_crs(insitu_shp)
# Print the CRS
print(crs_insitu)


crs_target <- st_crs("+init=EPSG:32645")

insitu_projected <- st_transform(insitu_shp, crs_target)


# Extract NDVI values for insitu data locations
ndvi_values <- raster::extract(ndvi, insitu_projected)

# Add the extracted NDVI values to the insitu_filtered_projected data
insitu_combined <- cbind(insitu_projected, NDVI = ndvi_values)

head(insitu_combined)

nrow(insitu_combined)

# Remove rows with NA values in the NDVI column
insitu_combined <- insitu_combined[!is.na(insitu_combined$NDVI), ]
nrow(insitu_combined)
```

### Data Splitting and Random Forest Model Training

-   Defined the proportion for the training data (70% for training, 30%
    for testing).
-   Set a seed for reproducibility.
-   Initialized empty data frames for training and testing sets.
-   Defined class names: "Bamboo," "Orchid," "PaddyRice," "Sugarcane,"
    "OtherCrop," and "Mixed_Area."
-   Splitted the data into training and testing sets with at least one
    sample from each class.
-   For each class:
    -   Filtered rows for the current class.
    -   Calculated the number of samples for training.
    -   Randomly selected samples for training and the rest for testing.
    -   Add the selected samples to the training and testing sets.
    -   Check the composition of the training and testing sets.
-   Trained a random forest model. Converted the "LU_CT" variable to a
    factor.
-   Trained the random forest model with "LU_CT" predicted based on the
    "NDVI" variable.I created the model with 100 trees (ntree = 100).

```{r}
# Define the proportion for the training data (70% for training, 30% for testing)
train_prop <- 0.7

# Set a seed for reproducibility
set.seed(123)

# Initialize empty data frames for training and testing sets
train_data <- data.frame()
test_data <- data.frame()

# Define class names
class_names <- c("Bamboo", "Orchid", "PaddyRice", "Sugarcane", "OtherCrop", "Mixed_Area")

# Split the data into training and testing sets with at least one sample from each class
for (class_name in class_names) {
  # Filter rows for the current class
  class_data <- insitu_combined[insitu_combined$LU_CT == class_name, ]
  
  # Calculate the number of samples for training
  num_train_samples <- round(train_prop * nrow(class_data))
  
  # Randomly select samples for training and the rest for testing
  class_train_index <- sample(1:nrow(class_data), num_train_samples)
  class_test_index <- setdiff(1:nrow(class_data), class_train_index)
  
  # Add the selected samples to the training and testing sets
  train_data <- rbind(train_data, class_data[class_train_index, ])
  test_data <- rbind(test_data, class_data[class_test_index, ])
}

# Check the composition of the training and testing sets
table(train_data$LU_CT)
table(test_data$LU_CT)

# Train the random forest model
train_data$LU_CT <- as.factor(train_data$LU_CT)
rf_model <- randomForest(LU_CT ~ NDVI, data = train_data, ntree = 100)
```

###Model Prediction and Evaluation

-   Made predictions on the test data using the trained random forest
    model.
-   Created a confusion matrix to compare predicted and actual classes.
-   Printed the confusion matrix as well as the head map.

```{r}
# Make predictions on the test data
rf_predictions <- predict(rf_model, test_data)

predicted_classes <- predict(rf_model, test_data)
actual_classes <- test_data$LU_CT
confusion_matrix <- table(Predicted = predicted_classes, Actual = actual_classes)
print(confusion_matrix)


# Convert the confusion matrix to a data frame
confusion_matrix_df <- as.data.frame(as.table(confusion_matrix))

# Rename the columns for better labels
colnames(confusion_matrix_df) <- c("Predicted", "Actual", "Frequency")

# Create the heatmap using ggplot2
heatmap_plot <- ggplot(confusion_matrix_df, aes(Actual, Predicted, fill = Frequency)) +
  geom_tile() +
  geom_text(aes(label = Frequency), vjust = 1, size = 4) +
  scale_fill_gradient(low = "pink", high = "red") +  # Define color palette
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels
  labs(title = "Confusion Matrix Heatmap", x = "Actual", y = "Predicted")

# Display the heatmap
print(heatmap_plot)
```

### Calculation of Accuarcy, Precision, Recall, and F1-Score

-   Calculated the overall accuracy of the model using the confusion
    matrix and displayed it to two decimal places.

-   Calculated precision, recall, and F1-score values for each class
    using the confusion matrix and Stored these values in their
    respective vectors.

-   Created a data frame to organize the precision, recall, and F1-score
    results for each class.

-   Used the kable function to format the results as a table.

-   Displayed the precision, recall and F1-Score value in form of Bar
    Diagram

```{r}
# Calculate accuracy
accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
cat("Accuracy: ", round(accuracy, 2), "\n")


# Initialize vectors to store precision, recall, and F1 values
precision_values <- numeric(length(class_names))
recall_values <- numeric(length(class_names))
f1_values <- numeric(length(class_names))


for (i in 1:length(class_names)) {
  class_name <- class_names[i]
  TP <- confusion_matrix[class_name, class_name]
  FP <- sum(confusion_matrix[, class_name]) - TP
  FN <- sum(confusion_matrix[class_name, ]) - TP
  precision <- TP / (TP + FP)
  recall <- TP / (TP + FN)
  f1_value <- 2 * (precision * recall) / (precision + recall)
  
 
  
  # Store values in respective vectors
  precision_values[i] <- precision
  recall_values[i] <- recall
  f1_values[i] <- f1_value
}

# Create a data frame with  results
results_df <- data.frame(
  Class = class_names,
  Precision = round(precision_values, 2),
  Recall = round(recall_values, 2),
  F1_Score = round(f1_values, 2)
)

# Use kable to format the results as a table
kable(results_df, caption = "Precision, Recall, and F1-Score for Each Class")

# Transform the data for a grouped bar chart
results_long <- results_df %>%
  pivot_longer(cols = c(Precision, Recall, F1_Score), names_to = "Metric", values_to = "Value")

# Create a grouped bar chart for Precision, Recall, and F1-Score
grouped_bar <- ggplot(results_long, aes(Class, Value, fill = Metric)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Performance Metrics for Each Class", x = "Class", y = "Value") +
  scale_fill_manual(values = c("Precision" = "blue", "Recall" = "green", "F1_Score" = "red")) +
  theme_minimal() +
  theme(legend.title = element_blank())

# Display the grouped bar chart
print(grouped_bar)
```

### Calculate Producer Accuracy (PA) and User Accuracy (UA)

-   Created empty vectors to store producer accuracy (PA) and user
    accuracy (UA) values for each class.
-   Calculated Producer Accuracy and User Accuracy for each class by
    dividing the true positive count by the sum of the actual counts for
    that class.
-   Created Producer and User Accuracy Table and used the kable function
    to format the results as a table.
-   Displayed the Producer Accuracy and User Accuracy value in form of
    Bar Diagram

```{r}
# Initialize vectors to store producer accuracy (PA) and user accuracy (UA)
PA_values <- numeric(length(class_names))
UA_values <- numeric(length(class_names))


for (i in 1:length(class_names)) {
  class_name <- class_names[i]
  
  # Calculate Producer Accuracy (PA)
  PA <- confusion_matrix[class_name, class_name] / sum(confusion_matrix[class_name, ])
  
  # Calculate User Accuracy (UA)
  UA <- confusion_matrix[class_name, class_name] / sum(confusion_matrix[, class_name])
  

  
  # Store values in respective vectors
  PA_values[i] <- PA
  UA_values[i] <- UA
}

# Create a data frame for  Producer Accuracy and User Accuracy results
PA_UA_df <- data.frame(
  Class = class_names,
  Producer_Accuracy = round(PA_values, 2),
  User_Accuracy = round(UA_values, 2)
)

# Use kable to format the results as a table
kable(PA_UA_df, caption = "Producer and User Accuracy for Each Class")

# Transform the data for a grouped bar chart for PA and UA
PA_UA_long <- PA_UA_df %>%
  pivot_longer(cols = c(Producer_Accuracy, User_Accuracy), names_to = "Metric", values_to = "Value")

# Create a grouped bar chart for PA and UA
grouped_bar_PA_UA <- ggplot(PA_UA_long, aes(Class, Value, fill = Metric)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Producer and User Accuracy for Each Class", x = "Class", y = "Value") +
  scale_fill_manual(values = c("Producer_Accuracy" = "orange", "User_Accuracy" = "maroon")) +
  theme_minimal() +
  theme(legend.title = element_blank())

# Display the grouped bar chart for PA and UA
print(grouped_bar_PA_UA)


```
